---
title: "Correlation and experiment design"
author: "Anil Goyal"
date: "31/05/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

# Linear relationship/Correlation

- between -1 to 1

Examples

## Correlation

- `cor()`

- use argument `use = "pairwise.complete.obs"` to handle missing values in any of the variables

__Pearson Moment Product Correlation__

Mathematical formula

Others - Kendall's tau
- Spearman's rho


```{r}
iris %>% 
  ggplot(aes(Sepal.Length, Sepal.Width)) +
  geom_point() +
  geom_smooth(method = 'lm', formula = "y ~ x")
```

Correlation

```{r}
cor(iris$Sepal.Length, iris$Sepal.Width)
```

## Correlation caveats

Non-linear relationships

When data (of `x`) is skewed, we can apply some transformation to `x` variable and/or `y` variable

- log transformation
- sqrt transformation
- reciprocal transformation
- combination of these

- correlation does not imply causation (spurious correlation)
- hidden variables at play (confounder)
e.g. Sales may be higher on holidays but it is hard to tell how much of the increased sales is due to holidays, and how much is due to the special deals and promotions that often run around holidays. Here, special deals confound the relationship between holidays and sales.


## Design of Experiments

1. Design of experiments
Often, data is created as a result of a study that aims to answer a specific question. However, data needs to be analyzed and interpreted differently depending on how the data was generated and how the study was designed.

2. Vocabulary
Experiments generally aim to answer a question in the form, "What is the effect of the treatment on the response?" In this setting, treatment refers to the explanatory or independent variable, and response refers to the response or dependent variable. For example, what is the effect of an advertisement on the number of products purchased? In this case, the treatment is an advertisement, and the response is the number of products purchased.

3. Controlled experiments
In a controlled experiment, participants are randomly assigned to either the treatment group or the control group, where the treatment group receives the treatment and the control group does not. In our example, the treatment group will see an advertisement, and the control group will not. Other than this difference, the groups should be comparable so that we can determine if seeing an advertisement causes people to buy more. If the groups aren't comparable, this could lead to confounding, or bias. If the average age of participants in the treatment group is 25 and the average age of participants in the control group is 50, age could be a potential confounder if younger people are more likely to purchase more, and this will make the experiment biased towards the treatment.

4. The gold standard of experiments will use...
The gold standard, or ideal experiment, will eliminate as much bias as possible using certain tools. The first tool to help eliminate bias in controlled experiments is to use a randomized controlled trial. In a randomized controlled trial, participants are randomly assigned to the treatment or control group and their assignment isn't based on anything other than chance. Random assignment like this helps ensure that the groups are comparable. The second way is to use a placebo, which is something that resembles the treatment, but has no effect. This way, participants don't know if they're in the treatment or control group. This ensures that the effect of the treatment is due to the treatment itself, not the idea of getting the treatment. This is common in clinical trials that test the effectiveness of a drug. The control group will still be given a pill, but it's a sugar pill that has minimal effects on the response.

5. The gold standard of experiments will use...
In a double-blind experiment, the person administering the treatment or running the experiment also doesn't know whether they're administering the actual treatment or the placebo. This protects against bias in the response as well as the analysis of the results. These different tools all boil down to the same principle: if there are fewer opportunities for bias to creep into your experiment, the more reliably you can conclude whether the treatment affects the response.

6. Observational studies
The other kind of study we'll discuss is the observational study. In an observational study, participants are not randomly assigned to groups. Instead, participants assign themselves, usually based on pre-existing characteristics. This is useful for answering questions that aren't conducive to a controlled experiment. If you want to study the effect of smoking on cancer, you can't force people to start smoking. Similarly, if you want to study how past purchasing behavior affects whether someone will buy a product, you can't force people to have certain past purchasing behavior. Because assignment isn't random, there's no way to guarantee that the groups will be comparable in every aspect, so observational studies can't establish causation, only association. The effects of the treatment may be confounded by factors that got certain people into the control group and certain people into the treatment group. However, there are ways to control for confounders, which can help strengthen the reliability of conclusions about association.

7. Longitudinal vs. cross-sectional studies
The final important distinction to make is between longitudinal and cross-sectional studies. 

In a longitudinal study, the same participants are followed over a period of time to examine the effect of treatment on the response. 

In a cross-sectional study, data is collected from a single snapshot in time. 

If you wanted to investigate the effect of age on height, a cross-sectional study would measure the heights of people of different ages and compare them. However, the results will be confounded by birth year and lifestyle since it's possible that each generation is getting taller. 

In a longitudinal study,the same people would have their heights recorded at different points in their lives, so the confounding is eliminated. It's important to note that longitudinal studies are more expensive, and take longer to perform, while cross-sectional studies are cheaper, faster, and more convenient.



