---
title: "Benford Law"
author: "Anil Goyal"
date: "03/06/2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

# Digit Analysis Using Benford's Law

Natural guess

Background

Newcomb - 1881
Benford - 1938

Law
$P(d=d_i)=log(1 + 1/d_i)$ where $d_i = 1..9$

Plot
```{r echo=FALSE}
# Implement Benford's Law for first digit
benlaw <- function(d) log10(1 + 1 / d)

# Calculate expected frequency for d=5
benlaw(5)

# Create a dataframe of the 9 digits and their Benford's Law probabilities
df <- data.frame(digit = 1:9, probability = benlaw(1:9))

# Create barplot with expected frequencies
ggplot(df, aes(x = digit, y = probability)) + 
	geom_bar(stat = "identity", fill = "dodgerblue") + 
	xlab("First digit") + ylab("Expected frequency") + 
	scale_x_continuous(breaks = 1:9, labels = 1:9) + 
	ylim(0, 0.33) + theme(text = element_text(size = 11))
```

Benford showed it on 20 different datasets

Pickham later showed that law is invariant to scaling

Another example (to be modified)

```{r}
# Load package benford.analysis
library(benford.analysis)
data(census.2009)

# Check conformity
bfd.cen <- benford(census.2009$pop.2009, number.of.digits = 1) 
plot(bfd.cen, except = c("second order", "summation", "mantissa", "chi squared","abs diff", "ex summation", "Legend"), multiple = F) 

# Multiply the data by 3 and check conformity again
data <- census.2009$pop.2009 * 3
bfd.cen3 <- benford(data, number.of.digits=1)
plot(bfd.cen3, except = c("second order", "summation", "mantissa", "chi squared","abs diff", "ex summation", "Legend"), multiple = F)
```


## Fraud detection

Two digits

examples


Ref-

ISACA JOURNAL ARCHIVES - [Understanding and Applying Benford’s Law - 1 May 2011](https://www.isaca.org/resources/isaca-journal/past-issues/2011/understanding-and-applying-benfords-law)

# Detecting outliers

## Univariate

## multivariate
